#1_travel_time.Rmd#
=============
This file estimates the population-weighted average travel time from each Primary Health Network (PHN) to its nearest HIV-related facility in Australia, and examines its relationship with late HIV diagnoses.
Run this after 0-SetupProject.

Workflow (one stage per chunk):
- Initialise: load packages and set project folder paths.
- Input: read the HIV notification data, PHN boundaries, the Australia mask, and the four facility datasets (S100 prescribers, GPs, hospitals, sexual health clinics), then combine them into one set of points.
- Load runned travel time files: reload cached rasters and the per-PHN layer to skip the slow travel-time step. Run this OR the calculation chunk, not both.
- Travel time calculation: build the friction surface, compute least-cost travel time to the nearest facility, weight by the 2024 ABS population grid, summarise to PHN level, and cache the outputs.
- Travel time plotting and Logistic regression: produce the figures and fit the models relating PHN travel-time category to late diagnosis.

Facility sets (s100, s100_shc, all, gp) are selected in the calculation chunk. Cache paths, output/figure filenames, and the regression cut points must be kept consistent with the chosen set.

```{r Initialise}
##setwd()

#######################r initialization####################
# Clear workspace
rm(list=ls()) 
#install.packages(c("tidyverse","sf", "terra", "geodata", "readxl","tidyterra","gdistance","broom","exactextractr","ggnewscale"))
#install.packages("traveltime", repos = c("https://idem-lab.r-universe.dev"))
# Load libraries
# Install required packages if not already installed
library(tidyverse)   
library(sf)           
library(terra)
library(geodata)
library(traveltime)
library(readxl)
library(tidyterra)
library(gdistance)
library(broom)
library(exactextractr)
library(ggnewscale)

# Source to current directory and set working directory
basePath <- getwd()

# Restart R and set working directory to source file location. 

##r User inputs
##select project
selectedproject<- "late_hiv_diagnosis"

# Various directories
projectFolder <- file.path(basePath,"projects", selectedproject)
dataFolder <- file.path(basePath,"projects", selectedproject,"data")
resultsFolder <- file.path(basePath,"projects", selectedproject, "output")
figFolder <- file.path(basePath,"projects", selectedproject,"figures")

# NOTE: absolute path to the restricted-access source data on the analyst's
# machine. Not reproducible off this machine by design — the line-list is not
# public. Users: point HIVDataFolder to your own local copy obtained
# under the relevant data agreement. Nothing downstream writes raw data.
# HIV data folder - for sourcing latest notification cleaning files
HIVDataFolder <- file.path("C:", "onedrive", "OneDrive - UNSW", "HIV Trends Across Space and Time", 
  "General")

```


```{r Input}
# Specify project input files and load them
hivdataFile<- file.path(HIVDataFolder, "hiv_data_cleaned_20250731.csv")
# Read the CSV data
hiv_data_all <- read_csv(hivdataFile, col_types = cols(
  month_hiv_diag = col_character(),
  age_hiv_notif = col_double(),
  sex = col_character(),
  gender = col_character(),
  region_birth = col_character(),
  au_birth = col_character(),
  lang_home = col_character(),
  year_arrival_aus = col_double(),
  cd4_count_diag = col_double(),
  late_hiv_diag_flag = col_logical(), 
  newly_acq_hiv_flag = col_logical(),
  hiv_exposure_risk = col_character(),
  place_acq = col_character(),
  state_res = col_character(),
  phn_res = col_character(),
  phn_name = col_character(),
  previ_diag_overseas = col_logical(),
  id = col_double(),
  year_diag = col_double()
))

hiv_data_all <- hiv_data_all %>%
  mutate(
    advanced_hiv_diag_flag = case_when(
      # If late_hiv_diag_flag is NA, return NA
      is.na(late_hiv_diag_flag) ~ NA,
      # If both conditions are met (cd4 < 200 AND late diagnosis), return TRUE
      cd4_count_diag < 200 & late_hiv_diag_flag == TRUE ~ TRUE,
      # All other cases return FALSE
      TRUE ~ FALSE
    )
  )

hiv_data_all <- hiv_data_all %>%
  mutate(
    cd4_50_hiv_diag_flag = case_when(
      # If late_hiv_diag_flag is NA, return NA
      is.na(late_hiv_diag_flag) ~ NA,
      # If both conditions are met (cd4 < 50 AND late diagnosis), return TRUE
      cd4_count_diag < 50 & late_hiv_diag_flag == TRUE ~ TRUE,
      # All other cases return FALSE
      TRUE ~ FALSE
    )
  )

hiv_data <- hiv_data_all %>%
  filter(
    is.na(age_hiv_notif) | age_hiv_notif >= 18
  )

# Check data structure
str(hiv_data)

# Read PHN shapefile
shapefile_path <- file.path(dataFolder, "phn.shp")
phn_shp <- st_read(shapefile_path)

# Inspect shapefile attribute fields
names(phn_shp)

# Ensure phn_res matches the PHN shapefile ID
# Suppose PHN shapefile has a column named "PHN_CODE"
phn_data <- phn_shp %>%
  st_drop_geometry() %>%
  dplyr::select(PHN_CODE) # Adjust as needed

# Check matching PHNs
unique(hiv_data$phn_res)
unique(phn_data$PHN_CODE)

# Load Australia shapefile for masking
au_shapefile <- gadm(
 country = "Australia",
 level = 0,
 path = tempdir(),
 resolution = 2
 )
au_shapefile


# Define location dataset file paths
s100File   <- file.path(dataFolder, "s100_20250711.xlsx")
gp_path    <- file.path(dataFolder, "nhsd_general_practice_full_20250819.xlsx")
hosp_path  <- file.path(dataFolder, "nhsd_hospitals_full_20250819.xlsx")
shc_path   <- file.path(dataFolder, "public_sexual_health_clinics_20250818.xlsx")

# Process S100 data
s100_data <- read_excel(s100File) %>%
  dplyr::filter(label %in% c(0, 4) & version %in% c(1)) %>%  # Only keep label = 0 or 4 to avoid duplicates and version 1 for consistency
  dplyr::select(x, y) %>%
  dplyr::filter(!is.na(x), !is.na(y)) %>%
  distinct() %>%
  mutate(
    s100_label = 1,              # All S100 locations get s100_label = 1
    type = "others100"           # Type is "others100"
  )

# Process GP data
gp_data <- read_excel(gp_path) %>%
  filter(is.na(sexual_clinic) |sexual_clinic != 1) %>%     # Exclude sexual clinics to avoid duplicates with shc_path
  dplyr::select(lon_final, lat_final, s100_label) %>%
  rename(x = lon_final, y = lat_final) %>%
  dplyr::filter(!is.na(x), !is.na(y)) %>%
  distinct() %>%
  mutate(
    s100_label = case_when(
      s100_label == 1 ~ 1,
      is.na(s100_label) ~ 0,
      TRUE ~ 0
    ),
    type = "gp"
  )

# Process Hospital data
hosp_data <- read_excel(hosp_path) %>%
  filter(is.na(sexual_clinic) |sexual_clinic != 1) %>%     # Exclude sexual clinics to avoid duplicates with shc_path
  dplyr::select(lon_final, lat_final, s100_label) %>%
  rename(x = lon_final, y = lat_final) %>%
  dplyr::filter(!is.na(x), !is.na(y)) %>%
  distinct() %>%
  mutate(
    s100_label = case_when(
      s100_label == 1 ~ 1,
      is.na(s100_label) ~ 0,
      TRUE ~ 0
    ),
    type = "hospital"
  )

# Process Sexual Health Clinics data
shc_data <- read_excel(shc_path) %>%
  dplyr::filter(!is.na(x), !is.na(y)) %>% 
  #filter(free %in% c(1,2,5) ) %>% 
  # add filter for free clinics only
  dplyr::select(x, y, s100_label) %>%
  distinct() %>%
  mutate(
    s100_label = case_when(
      s100_label == 1 ~ 1,
      is.na(s100_label) ~ 0,
      TRUE ~ 0
    ),
    type = "shc"
  )

# Combine all datasets
location_combined_data <- bind_rows(
  s100_data,
  gp_data,
  hosp_data,
  shc_data
) %>%
  dplyr::select(x, y, s100_label, type) %>%
  distinct()  # Remove any remaining duplicates

location_combined_data %>%
  count(type, sort = TRUE)

location_combined_data %>%
  count(type, s100_label, sort = TRUE)

```

```{r load runned travel time files}
#### Load population raster with quick running####
# Reload cached outputs so the slow travel-time calculation can be skipped.
# Run EITHER this chunk OR the "Travel time calculation" chunk, never both.

# Paths
cacheFolder <- file.path(dataFolder, "_cache")
#s100 only
fric_file <- file.path(cacheFolder, "friction_motor2020_au_s100_20260414.tif")
tt_file   <- file.path(cacheFolder, "traveltime_au_motor2020_pointsset_v1_s100_20260414.tif")
pts_file  <- file.path(cacheFolder, "points_used_for_traveltime_s100_20260414.csv")
meta_file <- file.path(cacheFolder, "traveltime_au_metadata_s100_20260414.json")
phn_travel_file <- file.path(cacheFolder, "phn_travel_s100_20260414.gpkg")
#s100+shc
fric_file <- file.path(cacheFolder, "friction_motor2020_au_s100_shc_20260623.tif")
tt_file   <- file.path(cacheFolder, "traveltime_au_motor2020_pointsset_v1_s100_shc_20260623.tif")
pts_file  <- file.path(cacheFolder, "points_used_for_traveltime_s100_shc_20260623.csv")
meta_file <- file.path(cacheFolder, "traveltime_au_metadata_s100_shc_20260623.json")
phn_travel_file <- file.path(cacheFolder, "phn_travel_s100_shc_20260623.gpkg")
#all
fric_file <- file.path(cacheFolder, "friction_motor2020_au_all_20260414.tif")
tt_file   <- file.path(cacheFolder, "traveltime_au_motor2020_pointsset_v1_all_20260414.tif")
pts_file  <- file.path(cacheFolder, "points_used_for_traveltime_all_20260414.csv")
meta_file <- file.path(cacheFolder, "traveltime_au_metadata_all_20260414.json")
phn_travel_file <- file.path(cacheFolder, "phn_travel_all_20260623.gpkg")
#gp only
fric_file <- file.path(cacheFolder, "friction_motor2020_au_gp_20260603.tif")
tt_file   <- file.path(cacheFolder, "traveltime_au_motor2020_pointsset_v1_gp_20260603.tif")
pts_file  <- file.path(cacheFolder, "points_used_for_traveltime_gp_20260603.csv")
meta_file <- file.path(cacheFolder, "traveltime_au_metadata_gp_20260603.json")
phn_travel_file <- file.path(cacheFolder, "phn_travel_gp_20260603.gpkg")


# Reload the exact saved objects (lazy-open for rasters)
friction_au                 <- terra::rast(fric_file)
travel_time_au              <- terra::rast(tt_file)
selected_location_full_clean <- readr::read_csv(pts_file, show_col_types = FALSE)
meta                        <- jsonlite::read_json(meta_file, simplifyVector = TRUE)
phn_map_reach <- sf::st_read(phn_travel_file)


```


```{r Travel time calculation}
#### Jump this chunk if the above are reloaded ####
# WARNING: this chunk is slow and requires internet (friction surface download).
# calculate_travel_time() can take many minutes on the national grid.  
# Its purpose is to generate the cache that the reload chunk reads, normally run once.

# Choose ONE facility set. The second assignment overwrites the first — keep the
# line for the set you want and comment out the other. Whatever you pick here
# must match the cache filenames and the regression cut points downstream.
# Select locations for travel time calculation
# all
selected_location_full <- location_combined_data 
# s100 and shc
selected_location_full <- location_combined_data %>%
                           filter(s100_label == 1 | type ==  "shc")

selected_location_full %>%
  count(type, s100_label, sort = TRUE)

selected_location <- selected_location_full %>%
                     dplyr::select(x, y)

# 1. Get friction surface for Australia and mask to country boundary
friction_au <- get_friction_surface(
    surface = "motor2020",
    extent = au_shapefile
  )|> 
  mask(au_shapefile)


selected_location_sf <- st_as_sf(selected_location, coords = c("x","y"), crs = 4326)

# 2. Inspect the friction raster’s CRS and extent
crs(friction_au)        # e.g. "+proj=longlat +datum=WGS84…"
ext(rast(friction_au))  # get its bounding box

# 3. Reproject points to match
selected_location_proj <- st_transform(selected_location_sf, crs(friction_au))

# 4. filter to only those within the raster’s extent
bb <- ext(rast(friction_au))
coords <- st_coordinates(selected_location_proj)
inside <- coords[,1] >= bb$xmin & coords[,1] <= bb$xmax &
          coords[,2] >= bb$ymin & coords[,2] <= bb$ymax
selected_location_clean <- coords[inside, , drop=FALSE]
selected_location_sf_clean <- selected_location_proj[inside, , drop = FALSE]
selected_location_full_clean <- selected_location_full[inside, ]


# 5. Run travel time
travel_time_au <- calculate_travel_time(
  friction_surface = friction_au,
  points           = selected_location_clean
)

#### save the rasters ####
cacheFolder <- file.path(dataFolder, "_cache")

fric_file <- file.path(cacheFolder, "friction_motor2020_au_s100_shc_20260707.tif")
tt_file   <- file.path(cacheFolder, "traveltime_au_motor2020_pointsset_v1_s100_shc_20260707.tif")
pts_file  <- file.path(cacheFolder, "points_used_for_traveltime_s100_shc_20260707.csv")
meta_file <- file.path(cacheFolder, "traveltime_au_metadata_s100_shc_20260707.json")

# Write friction raster only if not already present
if (!file.exists(fric_file)) {
  terra::writeRaster(
    friction_au, fric_file,
    overwrite = FALSE,
    gdal = c("TILED=YES", "COMPRESS=LZW", "BIGTIFF=YES")
  )
} else {
  message("File already exists: ", fric_file)
}

# Write points CSV only if not already present
if (!file.exists(pts_file)) {
  readr::write_csv(as.data.frame(selected_location_full_clean), pts_file)
} else {
  message("File already exists: ", pts_file)
}

# Write travel time raster only if not already present
if (!file.exists(tt_file)) {
  terra::writeRaster(
    travel_time_au, tt_file,
    overwrite = FALSE,
    datatype = "FLT4S",
    NAflag = -9999,
    gdal = c("TILED=YES", "COMPRESS=LZW", "BIGTIFF=YES")
  )
} else {
  message("File already exists: ", tt_file)
}

# Write metadata only if not already present
if (!file.exists(meta_file)) {
  meta <- list(
    created_at = format(Sys.time(), "%Y-%m-%d %H:%M:%S"),
    surface = "motor2020",
    raster_file = tt_file,
    friction_file = fric_file,
    n_points = nrow(selected_location_clean),
    crs_wkt = as.character(terra::crs(travel_time_au))
  )
  jsonlite::write_json(meta, meta_file, pretty = TRUE, auto_unbox = TRUE)
} else {
  message("File already exists: ", meta_file)
}


#### Travel time of PHN based on 2024 population grid ~1km ####
# Population weighting uses the 2024 ABS 1-km grid. Units: travel time in MINUTES,
# population as PERSON COUNTS per cell (density is reconstructed then re-aggregated
# to the travel-time grid to preserve totals under reprojection).

# 1-km ABS population grid (density: people per km^2)
Sys.setenv(PROJ_NETWORK = "OFF")

pop_grid <- terra::rast(file.path(dataFolder, "abs_popgrid_1km_2024.tif"))


# Source (counts/cell) -> density (people/km^2) on the *source* grid
area_src <- terra::cellSize(pop_grid, unit = "km")
dens_src <- pop_grid / area_src

# Project density to the *target* 
# Make sure travel_time_au has a valid CRS first:
terra::crs(travel_time_au) <- terra::crs(friction_au)
dens_tgt <- terra::project(dens_src, travel_time_au, method = "bilinear")

# Convert density back to counts on the *target* grid
area_tgt <- terra::cellSize(travel_time_au, unit = "km")
pop_count_res <- dens_tgt * area_tgt

# Mask to Australia
pop_count_res <- terra::mask(pop_count_res, au_shapefile)

# Clean inputs
travel_time_au_clean <- terra::ifel(is.finite(travel_time_au), travel_time_au, NA)


#### Within-cell travel time for facility cells (replace tt == 0) ####
# The least-cost surface snaps each facility to its cell centroid and assigns
# that cell travel time = 0, implying residents sharing a cell with a facility
# have zero travel. We replace that 0 with the EXPECTED within-cell travel time:
# the mean straight-line distance between two random points in the cell, turned
# into minutes using the cell's own friction value.

# Expected distance between two uniform random points in a unit square
# (Robbins constant ≈ 0.5214). Used to estimate mean within-cell travel for
# cells that contain a facility (assigned travel time 0 by the least-cost surface).
square_random <- (2 + sqrt(2) + 5 * asinh(1)) / 15            # = 0.5214054

# Keep friction on the same grid/CRS as the travel-time raster
terra::crs(travel_time_au_clean) <- terra::crs(friction_au)
if (!terra::compareGeom(friction_au, travel_time_au_clean, stopOnError = FALSE)) {
  friction_au <- terra::resample(friction_au, travel_time_au_clean, method = "bilinear")
}

# Per-cell effective side length (m): sqrt of the true cell area
# E-W shrinkage with latitude is handled per cell
eff_side_m <- sqrt(terra::cellSize(travel_time_au_clean, unit = "m"))   # m

# Expected within-cell distance (m), then minutes via friction (minutes/metre)
within_cell_dist_m <- square_random * eff_side_m
circuity           <- 1            
within_cell_time   <- within_cell_dist_m * friction_au * circuity   # minutes

# Patch ONLY the facility cells (exactly 0 in the accCost output); leave the rest

travel_time_au_corr <- terra::ifel(travel_time_au_clean == 0, within_cell_time, travel_time_au_clean)
names(travel_time_au_corr) <- "travel_time"

# Sanity check on the patched cells
pv <- terra::values(within_cell_time)[terra::values(travel_time_au_clean) == 0]
pv <- pv[is.finite(pv)]
message("Facility cells patched: ", length(pv))
print(round(quantile(pv, c(0,.05,.25, .5,.75, .9, 1), na.rm = TRUE), 3))   # minutes

#### Calculate population-weighted average travel time for each PHN ####
pop_count_res        <- terra::ifel(is.finite(pop_count_res) & pop_count_res >= 0, pop_count_res, NA)

# Stack so the reducer sees both layers
stack_tt_pop <- c(travel_time_au_corr, pop_count_res)

phn_avg_reach <- exact_extract(
  x = stack_tt_pop,
  y = phn_shp,
  fun = function(df, cf) {
    # df[[1]] = travel time; df[[2]] = pop count per cell
    tt  <- df[[1]]
    pop <- df[[2]]

    keep <- is.finite(tt) & is.finite(pop) & (pop > 0) &
            is.finite(cf) & (cf > 0)

    if (!any(keep)) return(NA_real_)  # no reachable population -> NA
    sum(tt[keep] * pop[keep] * cf[keep]) / sum(pop[keep] * cf[keep])
  },
  progress = TRUE
)

phn_map_reach <- phn_shp |>
  dplyr::mutate(avg_drive_min_reach = phn_avg_reach)


phn_travel_file <- file.path(cacheFolder, "phn_travel_all_20260707.gpkg")

if (!file.exists(phn_travel_file)) {
  sf::st_write(
    phn_map_reach,
    dsn = phn_travel_file,
    driver = "GPKG"
  )
  message("✅ Saved new file: ", phn_travel_file)
} else {
  message("⚠️ File already exists, skipping: ", phn_travel_file)
}

phn_travel_csv <- file.path(resultsFolder, "phn_travel_all_20260707.csv")

if (!file.exists(phn_travel_csv)) {
  phn_map_reach |>
    sf::st_drop_geometry() |>
    dplyr::select(PHN_NAME, avg_drive_min_reach) |>   # adjust name col if needed
    write.csv(file = phn_travel_csv, row.names = FALSE)
  message("✅ Saved new file: ", phn_travel_csv)
} else {
  message("⚠️ File already exists, skipping: ", phn_travel_csv)
}


#### Calculate proportion of people within X minutes ####
# valid populated cells
keep <- is.finite(values(travel_time_au_corr)) &
        is.finite(values(pop_count_res)) &
        values(pop_count_res) > 0

tt_vals  <- values(travel_time_au_corr)[keep]
pop_vals <- values(pop_count_res)[keep]

# proportion within X minutes
prop_within_5 <- sum(pop_vals[tt_vals <= 30]) / sum(pop_vals)

# percentage
pct_within_5 <- prop_within_5 * 100

pct_within_5


```



```{r travel time plotting}
# Figures reflect whichever facility set was reloaded/calculated above. The PHN
# map's colour scale (option, breaks) is tuned per set — see inline comments.

aus_bbox <- sf::st_bbox(sf::st_as_sf(phn_shp))

x_range <- c(aus_bbox["xmin"], aus_bbox["xmax"])
y_range <- c(aus_bbox["ymin"], aus_bbox["ymax"])

#### prescriber locations over friction surface ####
selected_location_mapping <- ggplot() +
  geom_spatraster(
    data = friction_au
  ) +
  geom_spatvector(
    data = au_shapefile,
    fill = "transparent",
    col = "grey50"
  ) +
  geom_point(
    data = selected_location_full_clean %>%  
      mutate(s100_label = as.character(s100_label)),
    aes(
      x = x,
      y = y,
      color = s100_label
    ),
    size = 0.5
  ) +
  scale_fill_viridis_c(
    option = "A",
    na.value = "transparent",
    direction = -1
  ) +
  scale_color_manual(
    values = c("0" = "#42d4f4", "1" = "#228B22"),  # Customize colors
    name = "S100 Prescriber",
    labels = c("0" = "No", "1" = "Yes")
  ) +
  labs(
    fill = "Resistance",
    x = element_blank(),
    y = element_blank()
  ) +
  theme_minimal()
selected_location_mapping

# City labels and leader lines

# 1. City anchor points
city_labels <- tribble(
  ~city,       ~lon,      ~lat,       ~dir,
  "Sydney",    151.2093,  -33.8688,   "R",
  "Adelaide",  138.6007,  -34.9285,   "DL",
  "Melbourne", 144.9631,  -37.8136,   "DL",
  "Perth",     115.8605,  -31.9505,   "L",
  "Canberra",  149.1300,  -35.2809,   "R",
  "Brisbane",  153.0251,  -27.4698,   "R",
  "Darwin",    130.8456,  -12.4634,   "L",
  "Hobart",    147.3272,  -42.8821,   "R"
)

# fine tuning
city_nudge <- tribble(
  ~city,       ~dx,    ~dy,
  "Sydney",     0.4,   0.2,
  "Adelaide",  -0.3,   0.0,
  "Melbourne", -0.3,  -0.1,
  "Perth",     -0.4,   0.0,
  "Canberra",   0.3,  -0.1,
  "Brisbane",   0.4,   0.2,
  "Darwin",     0.0,   0.0,
  "Hobart",     0.0,   0.0
)

city_labels <- city_labels %>%
  left_join(city_nudge, by = "city") %>%
  mutate(
    dx = replace_na(dx, 0),
    dy = replace_na(dy, 0)
  )

# 2. Direction vectors
bbox <- st_bbox(phn_map_reach)
x_span <- as.numeric(bbox["xmax"] - bbox["xmin"])
y_span <- as.numeric(bbox["ymax"] - bbox["ymin"])

dir_step <- list(
  U  = c(0.00, 0.030),
  D  = c(0.00,-0.030),
  L  = c(-0.025, 0.00),
  R  = c(0.020, 0.00),
  UL = c(-0.012, 0.010),
  UR = c(0.012, 0.010),
  DL = c(-0.025,-0.015),
  DR = c(0.025,-0.010)
)

get_dir_vectors <- function(x_span, y_span, dir_step) {
  lapply(dir_step, function(v) c(v[1] * x_span, v[2] * y_span))
}

dir_vec <- get_dir_vectors(x_span, y_span, dir_step)

# 3. Label positions
city_labels <- city_labels %>%
  rowwise() %>%
  mutate(
    x = lon,
    y = lat,
    x_lab = x + dir_vec[[dir]][1] + dx,
    y_lab = y + dir_vec[[dir]][2] + dy
  ) %>%
  ungroup()

# 4. Text alignment + line target at text edge
city_labels <- city_labels %>%
  mutate(
    hjust = case_when(
      dir %in% c("R", "UR", "DR") ~ 0,   # label extends to right
      dir %in% c("L", "UL", "DL") ~ 1,   # label extends to left
      TRUE ~ 0.5
    )
  )

# crude estimate of text width in map units
char_width <- 0.0035 * x_span

city_labels <- city_labels %>%
  mutate(
    text_width = str_length(city) * char_width,
    
    # where the leader line should point:
    # if label is on right and left-aligned, point to left edge
    # if label is on left and right-aligned, point to right edge
    x_text_edge = case_when(
      hjust == 0   ~ x_lab,
      hjust == 1   ~ x_lab,
      TRUE         ~ x_lab
    ),
    y_text_edge = y_lab
  )

# shorten slightly so line stops just before touching text
gap <- 0.002 * x_span

city_labels <- city_labels %>%
  mutate(
    x_line_end = case_when(
      dir %in% c("R", "UR", "DR") ~ x_text_edge - gap,
      dir %in% c("L", "UL", "DL") ~ x_text_edge + gap,
      TRUE ~ x_text_edge
    ),
    y_line_end = y_text_edge
  )

#### Plot PHN average travel time ####
phn_traveltime_mapping <- ggplot(phn_map_reach) +
  geom_sf(aes(fill = avg_drive_min_reach), 
          colour = "grey50", size = 0.2) +
  scale_fill_viridis_c(
  option    = "G",          # G for s100, D for all
  direction = -1,
#  transform = "log10",      # for s100
  na.value  = "transparent",
  breaks    = c(1, 5, 10),
  labels    = c(1, 5, 10),
#  breaks    = c(2, 5, 10, 30, 60, 120, 180), # for s100
#  labels    = c(2, 5, 10, 30, 60, 120, 180),# for s100
  name      = "Minutes"
) +
   coord_sf(
    xlim = x_range,
    ylim = y_range,
    expand = FALSE
  ) +
      geom_segment(
    data = city_labels,
    aes(x = x, y = y, xend = x_line_end, yend = y_line_end),
    linewidth = 0.5,
    color = "black"
  ) +
  geom_text(
    data = city_labels,
    aes(x = x_lab, y = y_lab, label = city, hjust = hjust),
    size = 3,
    color = "black",
    fontface = "bold"
  ) +

  theme_minimal() +
  labs(
    title    = "",#Population-Weighted Mean Travel Time by PHN
    subtitle = "",
    x        = NULL, 
    y        = NULL
  ) +
  theme(
    axis.text      = element_blank(),
    panel.grid     = element_blank(),
    plot.title     = element_text(face = "bold", size = 14),
    plot.subtitle  = element_text(size = 10)
  )
phn_traveltime_mapping

#### Plot travel time with prescriber locations each cell####


smallest_section_drivetime_mapping <- ggplot() +
  # Grey background for zero-pop cells
#  geom_spatraster(data = pop_zero_mask) +
#  scale_fill_gradientn(
#    colours = c("grey85","grey85"),
#    limits  = c(0, 1),
#    breaks  = 1,
#    labels  = "Population = 0",
#    na.value = NA,
#    name    = NULL
#  ) +
  ggnewscale::new_scale_fill() +
  
  # Travel time only where pop > 0 : use data = tt_nonzero
  geom_spatraster(data = travel_time_au) +
  scale_fill_viridis_c(
    option    = "A",
    direction = -1,
    limits    = c(0, 2500),
    breaks    = c(0, 500, 1000, 1500, 2000, 2500),
    labels    = c("0", "500", "1000", "1500", "2000", ">2500"),
    oob       = scales::squish,
    na.value  = "transparent",
    name      = "Minutes",
    guide     = guide_colorbar(order = 1)
  ) +
  
  # Prescriber points
  geom_point(
    data = selected_location_full_clean %>%
      dplyr::filter(!is.na(s100_label)) %>%
      dplyr::mutate(s100_label = as.character(s100_label)),
    aes(x = x, y = y, color = s100_label),
    size = 0.5
  ) +
  scale_color_manual(
    values = c("0" = "#42d4f4", "1" = "#228B22"),
    name   = "S100 Prescriber",
    labels = c("0" = "No", "1" = "Yes"),
    guide  = guide_legend(order = 2)
  ) +
  
  # PHN boundaries 
  geom_spatvector(
    data = phn_shp,
    fill = "transparent",
    col  = "grey50",
    linewidth = 0.2
  ) +
   coord_sf(
    xlim = x_range,
    ylim = y_range,
    expand = FALSE
  ) +
    geom_segment(
    data = city_labels,
    aes(x = x, y = y, xend = x_line_end, yend = y_line_end),
    linewidth = 0.5,
    color = "black"
  ) +
  geom_text(
    data = city_labels,
    aes(x = x_lab, y = y_lab, label = city, hjust = hjust),
    size = 3,
    color = "black",
    fontface = "bold"
  ) +
  theme_minimal() +
  labs(
    title    = "",# "Travel Time to Nearest Prescriber (Populated Cells Only)"
    subtitle = "", # Australia, motorised friction (2020); zero-population cells in grey
    x = NULL, y = NULL
  ) +
  theme(
    axis.text     = element_blank(),
    panel.grid    = element_blank(),
    plot.title    = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10)
  )

smallest_section_drivetime_mapping

####Save output figures####
SaveFigure <- function(folder, filename, figure, 
    format = ".png", width = 15, 
    height = 10, units = "cm", dpi = 300,...) {
    ggsave(file.path(folder, paste(filename, format, sep = "")), 
      plot = figure, width = width, height = height, units = units,dpi=dpi, ...)
}


if (!file.exists(file.path(figFolder, "grid_all_20260707.png"))) {
    SaveFigure(figFolder,"grid_all_20260707", selected_location_mapping)
}

if (!file.exists(file.path(figFolder, "grid_traveltime_all_20260707.png"))) {
    SaveFigure(figFolder,"grid_traveltime_all_20260707", smallest_section_drivetime_mapping)
}

if (!file.exists(file.path(figFolder, "phn_traveltime_all_20260707.png"))) {
    SaveFigure(figFolder,"phn_traveltime_all_20260707", phn_traveltime_mapping)
}


####conceptual figure for mean distance between two random points in a unit square####
set.seed(42)

## Palette
col_orig  <- "#E8833A"; col_fac <- "#2E6FB5"; col_add <- "#D1495B"
col_red   <- "#FBE7D3"; col_blue <- "#D3E4F5"; col_green <- "#CDE8C9"

## Grid: cell centres at integer coordinates, extent = centre +/- 0.5
cells <- expand.grid(cx = 1:5, cy = 1:4)
cells$xmin <- cells$cx - 0.5; cells$xmax <- cells$cx + 0.5
cells$ymin <- cells$cy - 0.5; cells$ymax <- cells$cy + 0.5
cells$fill <- "white"
cells$fill[cells$cx == 2 & cells$cy == 1] <- col_red
cells$fill[cells$cx == 2 & cells$cy == 4] <- col_blue
cells$fill[cells$cx == 5 & cells$cy == 3] <- col_green

centres <- data.frame(x = c(2, 2, 2, 2), y = c(1, 2, 3, 4))
res <- data.frame(x = 5 + runif(1, -0.35, -0.05), y = 3 + runif(1,  0.05, 0.35))
fac <- data.frame(x = 5 + runif(1,  0.05,  0.35), y = 3 + runif(1, -0.35, -0.05))

conceptual <- ggplot() +
  geom_rect(data = cells,
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = fill),
            colour = "grey80") +
  scale_fill_identity() +
  ## --- 1 km scale arrow above the top-left cell (cx=1, cy=4) ---
  geom_segment(aes(x = 0.5, xend = 1.5, y = 4.75, yend = 4.75),
               arrow = arrow(ends = "both", length = unit(0.12, "cm")),
               colour = "black", linewidth = 0.5) +
  annotate("text", x = 1, y = 5.0, label = "1 km",
           colour = "black", size = 3) +
  ## --- 1 km scale arrow left of the top-left cell (cx=1, cy=4) ---
  geom_segment(aes(x = 0.25, xend = 0.25, y = 3.5, yend = 4.5),
               arrow = arrow(ends = "both", length = unit(0.12, "cm")),
               colour = "black", linewidth = 0.5) +
  annotate("text", x = 0.05, y = 4.0, label = "1 km",
           colour = "black", size = 3, angle = 90) +
  ## --- cell-to-cell: map colour & linetype to a label for the legend ---
  geom_segment(aes(x = 2, y = 1, xend = 2, yend = 4,
                   colour = "Cell-to-cell", linetype = "Cell-to-cell"),
               linewidth = 0.7) +
  geom_point(data = centres, aes(x, y), colour = "black", size = 3) +
  annotate("text", x = 1.75, y = 0.75, label = "resident",
           colour = "#E8833A", fontface = 2, size = 3, hjust = 0) +
  annotate("text", x = 1.75, y = 4.25, label = "facility",
           colour = "#2E6FB5", fontface = 2, size = 3, hjust = 0) +

  ## --- within-cell: same trick with a different label ---
  geom_segment(aes(x = res$x, y = res$y, xend = fac$x, yend = fac$y,
                   colour = "Within-cell", linetype = "Within-cell"),
               linewidth = 1) +
  geom_point(data = res, aes(x, y), shape = 21, fill = col_orig,
             colour = "white", size = 4) +
  geom_point(data = fac, aes(x, y), shape = 23, fill = col_fac,
             colour = "white", size = 4) +
  annotate("text", x = res$x + 0.05, y = res$y + 0.25, label = "resident",
           colour = col_orig, fontface = 2, size = 3) +
  annotate("text", x = fac$x, y = fac$y - 0.18, label = "facility",
           colour = col_fac, fontface = 2, size = 3) +

  ## --- manual scales: identical name => merged into one legend ---
  scale_colour_manual(name = NULL,
    values = c("Cell-to-cell" = "grey40", "Within-cell" = col_add)) +
  scale_linetype_manual(name = NULL,
    values = c("Cell-to-cell" = "dashed", "Within-cell" = "solid")) +

  coord_equal(expand = FALSE, clip = "off") +
  labs(title = "") +
  theme_void(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 10, margin = margin(b = 4)),
        plot.margin = margin(6, 6, 6, 6),
        legend.position = "bottom")
conceptual

####Save output figures####
SaveFigure <- function(folder, filename, figure, 
    format = ".png", width = 15, 
    height = 10, units = "cm", dpi = 600,...) {
    ggsave(file.path(folder, paste(filename, format, sep = "")), 
      plot = figure, width = width, height = height, units = units,dpi=dpi, ...)
}



if (!file.exists(file.path(figFolder, "conceptual_20260707.png"))) {
    SaveFigure(figFolder,"conceptual_20260707", conceptual)
}



```



```{r logistic regression}
# Outcome: late HIV diagnosis (late_hiv_diag_flag). Analysis restricted to adults
# (>=18), diagnosed in Australia (previ_diag_overseas == FALSE), 2005-2024.
# Exposure of interest: PHN population-weighted travel-time category, interacted
# with diagnosis period.

#### Handle data ####

# Add age_group and merge in avg_drive_min
hiv_data_filter <- hiv_data %>%
  # a) age groups
  mutate(
    age_group = case_when(
      is.na(age_hiv_notif)          ~ NA_character_,
      age_hiv_notif < 30            ~ "<30",
      age_hiv_notif >= 30 & age_hiv_notif < 50  ~ "30-49",
      age_hiv_notif >= 50           ~ "50+"
    ),
    age_group = factor(age_group, levels = c("<30", "30-49", "50+"))
  ) %>%
  # b) join in PHN avg times by the PHN code column
  left_join(
    phn_map_reach %>% dplyr::select(PHN_CODE, avg_drive_min_reach_s100 = avg_drive_min_reach),
    by = c("phn_res" = "PHN_CODE")
  ) %>% 
  # c) travel‐time categories
  mutate(
    avg_travel_min_group_s100 = cut(
      avg_drive_min_reach_s100,
      breaks = c(-Inf, 2, 5, Inf),     #c(-Inf, 2, 5, Inf) for sensitivity analysis, c(-Inf,10,30, Inf) for main analysis
      labels = c("<2", "2-<5","5+"),  #c("<2", "2-<5", "5+") for sensitivity analysis, c("<10", "10-<30", "30+") for main analysis
      right  = FALSE
    ),
    avg_travel_min_group_s100 = factor(
      avg_travel_min_group_s100,
      c("<2", "2-<5","5+"),  #c("<2", "2-<5", "5+") for sensitivity analysis, c("<10", "10-<30", "30+") for main analysis
    )
  )

names(hiv_data_filter)

hiv_data_filter <- hiv_data_filter %>%
  mutate(
    # 1. Collapse into three groups + preserve NAs
    hiv_exposure_risk_group = case_when(
      hiv_exposure_risk %in% c("MSM", "MSMIDU") ~ "MSM",
      hiv_exposure_risk == "Hetero"            ~ "Hetero",
      is.na(hiv_exposure_risk)                  ~ NA_character_,
      TRUE                                      ~ "Other"
    ),
    # 2. Turn into a factor with MSM as the reference level
    hiv_exposure_risk_group = factor(
      hiv_exposure_risk_group,
      levels = c("MSM", "Hetero", "Other")
    ),
    # 3. Re-factor sex so that Male is the reference
    sex = factor(sex, levels = c("Male", "Female", "Other")))

# Inspect new vars
hiv_data_filter %>% 
  count(age_group, avg_travel_min_group_s100, sort = TRUE) %>% 
  print(n = 10)

#### Logistic regression ####
hiv_data_logistic <- hiv_data_filter %>%
  filter(
    previ_diag_overseas == FALSE,
    year_diag >= 2005,
    year_diag <= 2024
  ) %>%
  mutate(
    period2 = case_when(
      year_diag >= 2005 & year_diag <= 2014 ~ "2005-2014",
      year_diag >= 2015 & year_diag <= 2024 ~ "2015-2024",
      TRUE ~ NA_character_
    ),
    period2 = factor(period2, levels = c("2005-2014","2015-2024"))
  )

#overall data summary
vars <- c(
  "hiv_exposure_risk_group",
  "sex",
  "au_birth",
  "age_group",
  "avg_travel_min_group_s100",
  "period2"
)

group_summaries <- map_dfr(vars, function(v) {
  
  hiv_data_logistic %>%
    # drop NAs in the grouping var
    filter(!is.na(.data[[v]])) %>%
    group_by(level = .data[[v]]) %>%
    summarise(
      n       = n(),
      late_n  = sum(late_hiv_diag_flag, na.rm = TRUE),
      prop_late = late_n / n,
      .groups = "drop"
    ) %>%
    mutate(variable = v) %>%
    dplyr::select(variable, level, n, late_n, prop_late)
})

print(group_summaries)

# Univariate logistic regression for each variable

# Fit one glm per variable and tidy the results
univ_results <- map_dfr(vars, function(v) {
  # build formula
  fml <- as.formula(paste0("late_hiv_diag_flag ~ ", v))
  
  # fit univariate model
  m <- glm(fml, data = hiv_data_logistic, family = binomial(link = "logit"))
  
  # tidy: exponentiate to OR, get 95% CI and p-value
  tidy(m, exponentiate = TRUE, conf.int = TRUE) %>%
    # keep only the predictor row (drop intercept)
    filter(term != "(Intercept)") %>%
    # add a column for which variable this was
    mutate(variable = v) %>%
    # rename for clarity
    dplyr::select(variable, term, OR = estimate, `2.5 %` = conf.low, `97.5 %` = conf.high, p.value)
})

#
print(univ_results)

#### Fit multivariate logistic regression: late_hiv_diag_flag, advanced_hiv_diag_flag, cd4_50_hiv_diag_flag ####
model <- glm(
  late_hiv_diag_flag ~  sex  + age_group+hiv_exposure_risk_group +au_birth+avg_travel_min_group_s100* period2, 
  data   = hiv_data_logistic,
  family = binomial(link = "logit")
)
summary(model)

# Extract coefficients and calculate odds ratios
odds_ratios <- exp(coef(model))
# Get the 95% CI on the coefficient (log‐odds) scale
ci_logodds <- confint(model)

# Exponentiate both the point estimates and the CIs
or_ci <- exp(cbind(
  OR      = coef(model),
  CI_low  = ci_logodds[,1],
  CI_high = ci_logodds[,2]
))

# 3. label and view
colnames(or_ci) <- c("OR", "2.5 %", "97.5 %")
print(or_ci)

sumry  <- summary(model)
pvals  <- sumry$coefficients[, "Pr(>|z|)"]
or_ci_p <- cbind(
  or_ci,
  p_value = pvals
)


# If file not already created make sure it is not overwritten
if (!file.exists(file.path(resultsFolder, "paper_logistic_regression_periods1_20260707.csv"))) {
write.csv(or_ci_p, file.path(resultsFolder, "paper_logistic_regression_periods1_20260707.csv"), row.names=TRUE)
}

if (!file.exists(file.path(resultsFolder, "paper_unilogistic_regression_20260707.csv"))) {
write.csv(univ_results, file.path(resultsFolder, "paper_unilogistic_regression_20260707.csv"), row.names=TRUE)
}

if (!file.exists(file.path(resultsFolder, "paper_frequency_table_20260707.csv"))) {
write.csv(group_summaries, file.path(resultsFolder, "paper_frequency_table_20260707.csv"), row.names=TRUE)
}




```

